---
title: | 
  | **An absolute *dated* review with R** 
  | Apps, databases, calibrations and analysis
  | for radiocarbon and dendrochronological data
author: "Thomas Huet"
runtime: shiny
output: 
  html_document:
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 19, fig.height = 14)
library(kableExtra) 
library(dplyr)
library(knitr)
# library(magick)
# library(c14bazAAR)
# library(grDevices)
# library(sp)
# library(leaflet)
# library(Bchron)
mandat.fields <- c("labnr", "c14age", "c14std", "lon", "lat")
# GHimgs <- "https://github.com/zoometh/C14/tree/main/docs/imgs/"
# panel.map <- image_read("docs/imgs/panel_map.png")
# panel.map.coords <- image_read("docs/imgs/panel_map_coords.png")
# panel.calib <- image_read("docs/imgs/panel_calib.png")
```

Absolute dating, like [radiocarbon](#c14) and [dendrochronological](#dendro) data represent key axis to understand dynamics of past societies beyond local variations. Today, computer programming, web applications and open database, allow to share and analyze large amount of absolute chronological data 

# Radiocarbon data {#c14}

Today exists various web-based open access resources for radiocarbon [interactive management](#c14.app) and [database](#c14.db)

## Applications {#c14.app}

RShiny ([R](https://www.r-project.org/) + [Shiny](https://shiny.rstudio.com/)) allows to create interactive web forms for radiocarbon dates selection by geographical region of interest, absolute ages, cultural periods, quality of the date, etc. These apps offer also analysis tools like summing probability densities (SPD). 

```{r dat.app, echo=FALSE}
c14.app.names <- c("[**MedAfriCarbon**](https://theia.arch.cam.ac.uk/MedAfriCarbon/)",
                   "[**NeoNet**](https://zoometh.github.io/C14/neonet)",
                   "[**EUROEVOL_R**](https://neolithic.shinyapps.io/Euroevol_R/)",
                   "[**CARD 2.0**](https://www.canadianarchaeology.ca/)")
c14.app.geo <- c("South Mediterranean",
                 "Northern Central Mediterranean",
                 "Central, Western and Northern Europe",
                 "North America, Europe and Australia")
c14.app.chr <- c("~ 9,600 - 600 BC",
                 "~ 8,000 - 4,500 BC",
                 "~ 17,000 - 500 BC",
                 "~ 13,000 BC - 1500 AD (year 2011)")
c14.app.nbdat <- c("~ 1,600",
                   "~ 1,600",
                   "~ 14,000",
                   "~ 36,000 (year 2011)")
c14.app.df <- data.frame(name=c14.app.names,
                         loc=c14.app.geo,
                         chrono=c14.app.chr,
                         nb.14C=c14.app.nbdat,
                         stringsAsFactors = F)
kable(c14.app.df,"html") %>%
  kable_styling(full_width = FALSE, position = "center", font_size=14)
```

> <font size="2"> [MedAfriCarbon](https://theia.arch.cam.ac.uk/MedAfriCarbon/) offers an almost complete toolbox to manage radiocarbon dates. **NeoNet** and **EUROEVOL_R** share the same layout, mostly inherited from the EUROEVOL database, only the datasets are differents. The **NeoNet** database, an Excel Sheet. The **EUROEVOL_R** database is the an open source [EUROEVOL](http://discovery.ucl.ac.uk/1469811/) <br>... *visit the webpages of* [NeoNet](https://zoometh.github.io/C14/neonet/#app) *and* [EUROEVOL_R](https://zoometh.github.io/C14/euroevol)</font>

Programming language permits to handle radiocarbon data. [Shiny](https://shiny.rstudio.com/) and [Bchron](https://cran.r-project.org/web/packages/Bchron/vignettes/Bchron.html) easy to calibrate and plot radiocarbon dates

```{r calibrate, echo=TRUE}
library(Bchron)
library(shiny)
ui <- fluidPage(
  sidebarPanel(
    numericInput("bp", "BP", 5600),
    numericInput("sd", "+/-", 60)),
  mainPanel(
    plotOutput("plot", height=300))
)
server <- function(input, output) {
  output$plot <- renderPlot({
    c14date <- BchronCalibrate(ages=input$bp,
                               ageSds=input$sd,
                               calCurves='intcal13',
                               ids='date')
    plot(c14date)
  })
}
shinyApp(ui, server)
```

Beside [Bchron](https://cran.r-project.org/web/packages/Bchron/vignettes/Bchron.html), other R packages allows calibration and management of radiocarbon dates like  [rcarbon](https://cran.r-project.org/web/packages/rcarbon/index.html),  [OxcAAR](https://cran.r-project.org/web/packages/oxcAAR/index.html), [RChronoModel](https://cran.r-project.org/web/packages/RChronoModel/index.html), [stratigraphr](https://stratigraphr.joeroe.io/), [rbacon](https://cran.r-project.org/web/packages/rbacon/index.html), [ArchaeoPhases](https://cran.r-project.org/web/packages/ArchaeoPhases/index.html). The main dedicated apps for calibration are: [OxCal](https://c14.arch.ox.ac.uk/oxcal/), [CalPal](http://www.calpal-online.de/) and [Calib](http://calib.org/calib/), [BCal](https://bcal.shef.ac.uk/)

[c14bazAAR](https://github.com/ropensci/c14bazAAR) is a polyvalent package, allowing numerous operations. Using this package with [Leaflet](https://rstudio.github.io/leaflet/) allow to create interactive map of radiocarbon dates coming from different databases. 

> <font size="2">At first, we read the list of available database stored in a csv file on GitHub (`DBs.list.url <- 'https://raw.github.com...`), then we create a Shiny app with a selection menu for the databases (`selectInput("c14.db", ...`) and a reactive map (`output$map <- ...`)</font>

```{r echo = TRUE, warning=FALSE}
library(shiny)
library(c14bazAAR)
library(leaflet)
DBs.list.url <- 'https://raw.github.com/ropensci/c14bazAAR/master/data-raw/url_reference.csv'
DBs.list <- read.csv(DBs.list.url)
DBs <- DBs.list$db
ui <- fluidPage(selectInput("c14.db", label = "c14bazAAR c14 databases",
                            choices = DBs, selected = DBs[2], width = "25%"),
                leafletOutput("map", height=300)
)
server <- function(input, output) {
  output$map <- renderLeaflet({
    dbC14 <- get_c14data(input$c14.db)
    dbC14 <- dbC14[complete.cases(dbC14[, mandat.fields]), ]
    db.summary <- paste(as.character(nrow(dbC14)), "dates <br> from", 
                        as.character(max(dbC14$c14age)),
                        "<br> to",
                        as.character(min(dbC14$c14age)),"BP")
    dbC14$lbl <- paste0(dbC14$labnr, " [", dbC14$c14age, "BP +/- ", dbC14$c14std, "]")
    long.lat.clean <- dbC14
    leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
      addTiles() %>%
      addCircleMarkers(lng=dbC14$lon,
                       lat=dbC14$lat,
                       popup=dbC14$lbl,
                       radius = 0.5,
                       opacity = 0.3) %>%
      addControl(db.summary, position = "topright")
  })
}
shinyApp(ui, server)
```

R offers a comprehensive framework to calibrate, plot, sum and use temporal constraints -- stratigraphy, periodisations, etc., like bayesian analysis -- on radiocarbon datasets. 
Free software offering a dedicated solution for bayesian inference are: [ChronoModel](https://chronomodel.com/), [OpenBUGS](http://www.openbugs.net/w/FrontPage), [JAGS](http://mcmc-jags.sourceforge.net/), etc. For a overview of the available solutions, see the [R Bayesian task view](https://cran.r-project.org/web/views/Bayesian.html)

## Databases {#c14.db}

An almost exhaustive inventory of open databases is available on [c14bazAAR](https://github.com/ropensci/c14bazAAR#databases) and [telearchaeology](http://telearchaeology.org/c14-databses/). Some of these open database have form interface:

```{r dat.db, echo=FALSE}
c14.db.names <- c("[Radon](https://radon.ufg.uni-kiel.de/)",
                "[ORAU](https://c14.arch.ox.ac.uk)",
                "[14SEA Project](http://www.14sea.org/2_dates.html)",
                "[EX ORIENTE](https://www.exoriente.org/associated_projects/ppnd_sites.php)",
                "[BANADORA](http://www.archeometrie.mom.fr/banadora/)",
                "...")
c14.db.df <- data.frame(names=c14.db.names,
                        stringsAsFactors = F)
kable(c14.db.df,"html") %>%
  kable_styling(full_width = FALSE, position = "center", font_size=14)
```

But, other can also be simple download files: [download EUROEVOL](http://discovery.ucl.ac.uk/1469811), [download EUBAR](https://telearchaeology.org///EUBAR/EUBAR%20def.xlsx), [download "katsianis"](https://rdr.ucl.ac.uk/ndownloader/files/23166314), ...

# Dendrochronological data {#dendro}

As radiocarbon data, dendrochronological open database can be retrieve from the web and mapped. For example, the database of the British [Archaeology Data Service](https://archaeologydataservice.ac.uk/archives/view/vag_dendro/index.cfm) can be download and read  
  
> <font size="2"> Once read, the database need to be cleaned. After what, the original British National Grid coordinates are converted to lat/long coordinates with the packages [rnrfa](https://cran.r-project.org/web/packages/rnrfa/index.html) and [sf](https://cran.r-project.org/web/packages/sf/index.html) </font>

```{r echo = TRUE, warning=FALSE}
url.root <- 'https://archaeologydataservice.ac.uk/catalogue/adsdata/arch-282-11/'
dendro.list.url <- paste0(url.root, 'dissemination/csv/ind_ful_to50.csv&hs=true')
dendro.list <- read.csv(dendro.list.url, stringsAsFactors = F)
dendro <- dendro.list[nchar(dendro.list$GRID_LETTE) == 2, ]
dendro <- dendro[nchar(dendro$EASTING) == 3, ]
dendro <- dendro[nchar(dendro$NORTHING) == 3, ]
dendro <- dendro[!is.na(dendro$YEAR_SHORT), ]
dendro$GEO <- paste0(dendro$GRID_LETTE, dendro$EASTING, dendro$NORTHING)
dendro.coords <- rnrfa::osg_parse(dendro$GEO)
dendro$lon = dendro.coords[[1]]
dendro$lat = dendro.coords[[2]]
nDB <- dendro %>%
  sf::st_as_sf(coords = c("lon", "lat"), crs = 27700) %>%
  sf::st_transform(4326) %>%
  sf::st_coordinates() %>%
  as_tibble()
dendro.cln <- cbind(dendro, nDB)
dendro.sample <- head(dendro.cln[, c("LIST_SEQ", "LOCATION",
                                     "YEAR_SHORT", "PERIOD", "X", "Y")])
kable(dendro.sample, "html", row.names = F) %>%
  kable_styling(full_width = FALSE, position = "center", font_size=12)
```

[Shiny](https://shiny.rstudio.com/) and [Leaflet](https://rstudio.github.io/leaflet/) allow, for example, to select data filtered on a year threshold, for example
the year of the battle of Agincourt: 1415. 

> <font size="2">Once filtered, the dendrochronological data can be download with the download button </font>

```{r echo = TRUE, warning=FALSE}
library(shiny)
library(leaflet)
ui <- fluidPage(
  sidebarPanel(numericInput("dend.thrs", "Dates older than year:", 1415),
                downloadButton('downloadData', 'Download data')),
  mainPanel(leafletOutput("map", height = 600))
)
server <- function(input, output) {
  output$map <- renderLeaflet({
    dendro.dec <- dendro.cln[dendro.cln$YEAR_SHORT < input$dend.thrs, ]
    colpal <- colorBin("RdYlGn", dendro.dec$YEAR_SHORT, bins  = 10)
    dendro.dec$lbl <- paste("numero:", dendro.dec$LIST_SEQ, '<br>',
                            "year:", dendro.dec$YEAR_SHORT)
    leaflet() %>%
      addProviderTiles(providers$Stamen.Toner) %>%
      addCircleMarkers(data = dendro.dec,
                       lng = ~X,
                       lat = ~Y,
                       popup = ~lbl,
                       color = ~colpal(YEAR_SHORT),
                       radius = 0.3,
                       opacity = 1) %>%
      addLegend('topright', pal = colpal, values = dendro.dec$YEAR_SHORT,
                title = 'Calendar Year <br> Ranges',
                opacity = 1)
  })
  output$downloadData <- downloadHandler(
    filename = function() { 
      paste("dendro_before_year_", as.character(input$dend.thrs), ".csv", sep="")
    },
    content = function(file) {
      write.csv(dendro.dec, file)
    })
}
shinyApp(ui, server, options = list(height = 600))
```

