---
title: | 
  | **A radiocarbon review with R**: 
  | Apps, databases, calibrations and analysis
author: "Thomas Huet"
runtime: shiny
output: 
  html_document:
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 19, fig.height = 14)
library(kableExtra) 
library(dplyr)
library(knitr)
library(magick)
library(c14bazAAR)
library(grDevices)
library(sp)
library(leaflet)
library(Bchron)
DBs <- c("emedyd", "eubar", "euroevol", "context", "katsianis", "medafricarbon", "radon")
mandat.fields <- c("labnr", "c14age", "c14std", "lon", "lat")
GHimgs <- "https://github.com/zoometh/C14/tree/main/docs/imgs/"
panel.map <- image_read("docs/imgs/panel_map.png")
panel.map.coords <- image_read("docs/imgs/panel_map_coords.png")
panel.calib <- image_read("docs/imgs/panel_calib.png")
```

Today exists various web-based open access resources for radiocarbon [interactive management](#c14.app) and [database](#c14.db)

# Applications {#c14.app}

RShiny ([R](https://www.r-project.org/) + [Shiny](https://shiny.rstudio.com/)) allows to create interactive web forms for radiocarbon dates selection by geographical region of interest, absolute ages, cultural periods, quality of the date, etc. These apps offer also analysis tools like summing probability densities (SPD). 

```{r dat.app, echo=FALSE}
c14.app.names <- c("[**MedAfriCarbon**](https://theia.arch.cam.ac.uk/MedAfriCarbon/)",
                   "[**NeoNet**](https://zoometh.github.io/C14/neonet)",
                   "[**EUROEVOL_R**](https://neolithic.shinyapps.io/Euroevol_R/)",
                   "[**CARD 2.0**](https://www.canadianarchaeology.ca/)")
c14.app.geo <- c("South Mediterranean",
                 "Northern Central Mediterranean",
                 "Central, Western and Northern Europe",
                 "North America, Europe and Australia")
c14.app.chr <- c("~ 9,600 - 600 BC",
                 "~ 8,000 - 4,500 BC",
                 "~ 17,000 - 500 BC",
                 "~ 13,000 BC - 1500 AD (year 2011)")
c14.app.nbdat <- c("~ 1,600",
                   "~ 1,600",
                   "~ 14,000",
                   "~ 36,000 (year 2011)")
c14.app.df <- data.frame(name=c14.app.names,
                         loc=c14.app.geo,
                         chrono=c14.app.chr,
                         nb.14C=c14.app.nbdat,
                         stringsAsFactors = F)
kable(c14.app.df,"html") %>%
  kable_styling(full_width = FALSE, position = "center", font_size=14)
```

> <font size="2"> [MedAfriCarbon](https://theia.arch.cam.ac.uk/MedAfriCarbon/) offers an almost complete toolbox to manage radiocarbon dates. **NeoNet** and **EUROEVOL_R** share the same layout, mostly inherited from the EUROEVOL database, only the datasets are differents. The **NeoNet** database, an Excel Sheet. The **EUROEVOL_R** database is the an open source [EUROEVOL](http://discovery.ucl.ac.uk/1469811/) <br> ... vistit the webpages of [NeoNet](https://zoometh.github.io/C14/neonet/#app) and [EUROEVOL_R](https://zoometh.github.io/C14/euroevol)</font>

Programming language permits to handle radiocarbon data. [Shiny](https://shiny.rstudio.com/)and [Bchron](https://cran.r-project.org/web/packages/Bchron/vignettes/Bchron.html) easy to calibrate and plot radiocarbon dates

```{r calibrate.values, echo=TRUE}
fluidPage(
  sidebarPanel(
    shiny::numericInput("bp", "BP", 5600),
    shiny::numericInput("sd", "+/-", 60),
    actionButton("go", "calibrate")),
  mainPanel(
    plotOutput("plot"))
)
output$plot <- shiny::renderPlot({
  c14date <- BchronCalibrate(ages=input$bp,
                             ageSds=input$sd,
                             calCurves='intcal13',
                             ids='date')
  plot(c14date)
})
```

Beside [Bchron](https://cran.r-project.org/web/packages/Bchron/vignettes/Bchron.html), other R packages allows calibration and management of radiocarbon dates like  [rcarbon](https://cran.r-project.org/web/packages/rcarbon/index.html) and  [OxcAAR](https://cran.r-project.org/web/packages/oxcAAR/index.html). The main dedicated apps for calibration are: [OxCal](https://c14.arch.ox.ac.uk/oxcal/),        [CalPal](http://www.calpal-online.de/) and [Calib](http://calib.org/calib/)
  
Using the packages [c14bazAAR](https://github.com/ropensci/c14bazAAR) and [Leaflet](https://rstudio.github.io/leaflet/) allow to create interactive map of radiocarbon dates coming from different databases:

```{r echo = TRUE, warning=FALSE}
shiny::selectInput("c14.db", label = "c14 databases sample",
                   choices = DBs, selected = "katsianis", width = "25%")
leaflet::renderLeaflet({
  dbC14 <- get_c14data(input$c14.db)
  dbC14 <- dbC14[complete.cases(dbC14[, mandat.fields]), ]
  db.summary <- paste(as.character(nrow(dbC14)), "dates <br> from", 
                      as.character(max(dbC14$c14age)),
                                   "<br> to",
                                   as.character(min(dbC14$c14age)),"BP")
  dbC14$lbl <- paste0(dbC14$labnr, " [", dbC14$c14age, "BP +/- ", dbC14$c14std, "]")
  long.lat.clean <- dbC14
  leaflet(options = leafletOptions(zoomControl = FALSE)) %>%
    addTiles() %>%
    addCircleMarkers(lng=dbC14$lon,
                     lat=dbC14$lat,
                     popup=dbC14$lbl,
                     radius = 0.5,
                     opacity = 0.3) %>%
    addControl(db.summary, position = "topright")
})
```

R offers a comprehensive framework to calibrate, plot, sum and use temporal constraints -- stratigraphy, periodisations, etc., like bayesian analysis -- on radiocarbon datasets. 
Free software offering a dedicated solution for bayesian inference are: [ChronoModel](https://chronomodel.com/), [OpenBUGS](http://www.openbugs.net/w/FrontPage), [JAGS](http://mcmc-jags.sourceforge.net/), etc. For a overview of the available solutions, see the [R Bayesian task view](https://cran.r-project.org/web/views/Bayesian.html)

# Databases {#c14.db}

An almost exhaustive inventory of open databases is avalaible on [c14bazAAR](https://github.com/ropensci/c14bazAAR#databases) and [telearchaeology](http://telearchaeology.org/c14-databses/). Some of these open database have form interface:

```{r dat.db, echo=FALSE}
c14.db.names <- c("[Radon](https://radon.ufg.uni-kiel.de/)",
                "[ORAU](https://c14.arch.ox.ac.uk)",
                "[14SEA Project](http://www.14sea.org/2_dates.html)",
                "[EX ORIENTE](https://www.exoriente.org/associated_projects/ppnd_sites.php)",
                "[BANADORA](http://www.archeometrie.mom.fr/banadora/)",
                "...")
c14.db.df <- data.frame(names=c14.db.names,
                        stringsAsFactors = F)
kable(c14.db.df,"html") %>%
  kable_styling(full_width = FALSE, position = "center", font_size=14)
download.url.img <- "https://loading.io/icon/19ma0i"
```

But, other can also be simple download files: [download EUROEVOL](http://discovery.ucl.ac.uk/1469811), [download EUBAR](https://telearchaeology.org///EUBAR/EUBAR%20def.xlsx), [download "katsianis"](https://rdr.ucl.ac.uk/ndownloader/files/23166314), ...

